Supplementary Materials to ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic''

### 2022-02-09 write-up/letters/round1/response-ref-1-v1.tex

\vspace{5mm}
We agree that the original manuscript was too focused on the scientific communication perspective although it did contain original derivations related to extending the statistical error decomposition of Meng (2018) to account for measurement error.  The revised manuscript now fits well within the mission of AOAS by focusing on novel methods and clear application to a current and important scientific problem. In particular, we use the decomposition to motivate new methodology that leverages auxiliary information to address selection bias. In Section 2, we now introduce the data streams that illustrate this methodology. First, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.  The Facebook survey provides auxiliary information that is used to construct selection propensities, which are then incorporated into an inverse-probability weighting method (introduced in Section 4.1) to estimate time-varying active infection rates. Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2 we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.  We extend the decompositions to the weighting and doubly robust setting which then guide our discussion of practical implications.
\vspace{5mm}

\end{itemize}
\newpage

{\bf Response to AE4}
\begin{itemize}
\item {\it I’ve read through it now, and think there is value to what they are doing.  It is not rocket science, but that is OK to me if there is applied impact. I think that the author is trying to substantively think deeply about the measurement error and selection bias in case counts, which I constantly encounter and see is a major issue in managing this pandemic.  But I would like to see more practical guidelines in the discussion.  I see some promise here but no guarantees.}

\vspace{5mm}

\vspace{5mm}
We agree that the original work was not rocket science. While infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.  By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.
\vspace{5mm}

To build practical guidelines, we first worked on accessing high quality data.
To this end, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.
\vspace{5mm}

This data motivated our practical proposal: use time-varying surveys such as the Facebook symptom survey to provide auxiliary information that allows us to construct our inverse-probability weights (introduced in Section 4.1) to estimate time-varying active infection rates. Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2, we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.
\vspace{5mm}

The main limitation of our proposed approach is the limited covariate information being currently reported with case counts.  Specifically, symptom status, contact with a COVID-19 positive individual, and other covariate information is either not collected or not reported.  This guides the suggested real-world improvements to data collection that will enable our proposed approach to combine survey statistical methods with infectious disease modeling to improve our understanding of active infection rates over time.
\vspace{5mm}

\end{itemize}
\newpage

{\bf Response to Referee}
\begin{itemize}
\item {\it I skimmed through the MS very quickly. Although it doesn’t seem like rocket science, and may also be making points that are very well known to infectious disease researchers, it may be that the author is making some good points and, more importantly, may be translating a message well known in one field to another field. If si, then the MS might be worthwhile from a science communication perspective.}

\vspace{5mm}
We agree that the original work was not rocket science. While infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.  By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.
\vspace{5mm}

We have now moved well beyond the science communication perspective.  In particular, we use the decomposition to motivate new methodology that leverages auxiliary information to address selection bias. In Section 2, we now introduce the data streams that illustrate this methodology. First, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.  The Facebook survey provides auxiliary information that is used to construct selection propensities, which are then incorporated into an inverse-probability weighting method (introduced in Section 4.1) to estimate time-varying active infection rates.  Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2, we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.  We extend the decompositions to the weighting and doubly robust setting which then guide our discussion of practical implications.
\vspace{5mm}

The main limitation of our proposed approach is the limited covariate information being currently reported with case counts.  Specifically, symptom status, contact with a COVID-19 positive individual, and other covariate information is either not collected or not reported.  This guides the suggested real-world improvements to data collection that will enable our proposed approach to combine survey statistical methods with infectious disease modeling to improve our understanding of active infection rates over time.
\vspace{5mm}

\end{itemize}

\end{letter}

\end{document}
